Bootstrapping Multi-atlas Hippocampal Segmentation with MAGeT

### 2013-03-01 paper.Rnw
Added changes to the paper to reflect the new figures

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Screening Dataset Demographics Table                                  %
%                                                                             %
<<ADNI1-screening-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID < 2000,   
               method = "reverse", test=FALSE, overall=TRUE)
latex(tab, file="", size="scriptsize", 
    caption="ADNI1 1.5T Screening dataset demographics",
    title="tab:ADN1_1.5T_Screening_Dataset_Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-69-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c("295","413","619","685","729","782","938","954",
                 "1018","1155","907","981","222","324","448","546",
                 "553","572","602","610","814","1341","675","681",
                 "731","1130","41","68","70","101","249","293","316",
                 "414","698","1206","1222","1339","751","1030","3",
                 "5","10","16","22","53","168","183","241","1205",
                 "328","1095","991","213","159","337","343","642",
                 "753","1109","1217","29","56","1188","1293","149",
                 "382","431","1202")
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               adni_69_RIDs, method = "reverse", test=FALSE, overall=TRUE)

@@

<<ADN1_XVal_prep, include=FALSE>>=

# all data from the validation trials
t_data       = read.csv(gzfile('data/a2a_ants/results_2013_01_03.csv.gz'))
t_data       = subset(t_data, batch=="batch_1" & atlases %in% c(1,3,5,7,9))
a_data       = t_data[,] #read.csv(gzfile('data/a2a_tracc/results_2013_01_03.csv.gz'))
diagnoses    = read.csv('data/a2a_diagnoses.csv')
t_data$se    = NULL
t_data$sn    = NULL
t_data$j     = NULL
a_data$se    = NULL
a_data$sn    = NULL
a_data$j     = NULL       
# add registration method to data frames (before joining)
t_data       = cbind(t_data, reg_method = "ANIMAL")
a_data       = cbind(a_data, reg_method = "ANTS")

@@

all_data = merge(ma, mb, by=c("timestamp", "atlases", "batch", "label", "reg_method", "subject"))
all_data$k_diff = all_data$k.x - all_data$k.y
method_labels <- list("Majority Vote" = "majvote", 
                      "NMI Vote" = "nmi", 
                      "Cross-correlation Vote" = "xcorr")
levels(all_data$method.y) <- method_labels
all_data       = merge(all_data, diagnoses)
@

\begin{figure}
<<ADN1-Xval-k_diff, cache=TRUE>>=
ggplot(all_data, aes(x=templates.y, y=k_diff)) + 
  facet_grid(reg_method~method.y) + 
  stat_summary(fun.y=mean,geom='bar',position="dodge",
               aes(y=k_diff,fill=as.factor(atlases))) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Difference in mean Kappa between voting method and multiatlas" ) +
  coord_cartesian(ylim=c(0,0.07), xlim=c(0.5,10.5))
@
  \caption{Difference in mean Kappa between MAGeT brain and multi-atlas}
  \label{ADNI1_cross_validation_MA_difference}
\end{figure}

\begin{figure}
<<ADN1-Xval-k_diff-ANTS-20_tmpls, cache=TRUE>>=
a_20_tmpls = subset(all_data, 
                      reg_method  == "ANTS" &
                      method.y    == "Majority Vote" &
                      templates.y == 20)
ggplot(a_20_tmpls, aes(x=factor(atlases))) +
  stat_summary(fun.y=mean,geom='point',size=3,colour="red",aes(y=k.x)) +
  stat_summary(fun.y=mean,geom='line',colour="red",aes(y=k.x,group=1)) +
  stat_summary(fun.y=mean,geom='point',size=3,colour="blue",aes(y=k.y)) +
  stat_summary(fun.y=mean,geom='line',colour="blue",aes(y=k.y,group=1)) +
  labs(x="Number of Atlases", y = "Kappa") +
  scale_colour_hue(name="Method") + 
  coord_cartesian(ylim=c(0.7,0.9))
  #theme(legend.justification=c(1,0), legend.position=c(1,0))
@
  \caption{Difference in mean Kappa between multi-atlas and MAGeT brain using ANTS, 20 templates}
  \label{}
\end{figure}

\begin{figure}
<<ADN1-Xval-k_diff-ANTS-20_tmpls, cache=TRUE>>=
a_20_tmpls = subset(all_data, 
                      reg_method  == "ANTS" &
                      method.y    == "Majority Vote" &
                      templates.y == 20 &
                      atlases %in% c(1,9))
template_library = read.csv('data/a2a_ants/template_library_composition.csv')
template_library$subject = NULL
a_20_tmpls = merge(a_20_tmpls, template_library,
                   by.x=c('timestamp', 'templates.y'), 
                   by.y=c('timestamp', 'templates'))
ggplot(a_20_tmpls, aes(y=k.y, x=templates_CN,colour=DX)) +  # why does k.x look the same?
  facet_grid( . ~ atlases) + 
  geom_smooth(method='lm',fullrange=F)

@@

<<ADNI-baseline-volumes-prep,echo=F,cache=T>>=
yr1   = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
mb    = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_MAGeT_volumes.csv")
maper = read.csv("data/ADNI_baseline_volumes/MAPER_volumes.csv")
snt   = read.csv("data/ADNI_baseline_volumes/UCSFSNTVOL.csv")
ucsd  = read.csv("data/ADNI_baseline_volumes/UCSDVOL.csv")
fs    = read.csv("data/ADNI_baseline_volumes/UCSFFSX_02_15_12.csv")
fsl   = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_FSLFIRST_volumes.csv")

@@

\begin{figure}[h]
<<ADNI-baseline-volumes-boxplot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes-prep',cache=T>>=
totals = vols
qplot(DX,value,data=melt(totals,measure.vars=c("FS", "FSL","UCSD", "MAPER","MAGeT","SNT"),na.rm=T),
      colour=variable,geom="boxplot") + 
        xlab("Diagnosis") + ylab("Total Hippocampal Volume (mm3)")

@@

\begin{figure}[h]
<<ADNI-baseline-volumes-plot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes-boxplot',cache=T>>=
by_snt_volumes = melt(totals, measure.vars=c("FSL","MAPER","MAGeT", "FS", "UCSD"))
fsl=subset(by_snt_volumes, variable=="FSL")
maper=subset(by_snt_volumes, variable=="MAPER")
maget=subset(by_snt_volumes, variable=="MAGeT")

qplot(SNT,value,data=by_snt_volumes,
      colour=variable, geom="smooth", method="lm") + 
    geom_point() +
#     geom_text(aes(
#       label=paste("FSL       R=", round(cor(fsl$SNT,fsl$value, use="complete.obs"), 2), 
#                   ", p<", round(as.double(t.test(fsl$SNT,fsl$value, paired=T)["p.value"]), 2)),
#       x=1800,y=10000, hjust=0,size=2)) + 
#     geom_text(aes(
#       label=paste("MAPER R=", round(cor(maper$SNT,maper$value, use="complete.obs"), 2), 
#                   ", p<", t.test(maper$SNT,maper$value, paired=T)["p.value"]), 
#       x=1800,y= 9500, hjust=0, size=2)) +
#     geom_text(aes(
#       label=paste("MAGeT  R=", round(cor(maget$SNT,maget$value, use="complete.obs"),2), 
#                   ", p<", t.test(maget$SNT,maget$value, paired=T)["p.value"]), 
#       x=1800,y= 9000, hjust=0, size=2)) +
    xlab("Total Hippocampal Volume (Manual Segmentation)") + 
    ylab("Total Hippocampal Volume (Automatic Segmentation)")
@
  \caption{{\bf ADNI Baseline cohort.} Comparison of HC volumes by FreeSurfer
  (FSF), MAGeT brain (MAGeT), MAPER, and manual (SNT).}
  \label{ADNI-baseline-volumes-plot}
\end{figure}

@@

\begin{figure}
<<SZ_volumes,echo=F,fig=T, fig.align='center',cache=T>>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv")
qplot(Manual,MAGeT,data=SZ_volumes) + geom_smooth(method="lm") + 
#   geom_text(aes(
#       label=paste("R=", round(cor(Manual,MAGeT, use="complete.obs"), 2), 
#                   ", p<", t.test(Manual, MAGeT, paired=T)["p.value"]),
#       x=6500,y=10000, hjust=0,size=2)) +
    xlab("Total Hippocampal Volume (Manual Segmentation)") + 
    ylab("Total Hippocampal Volume (MAGeT Segmentation)")

@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total
  HC volumes for MAGeT against manually rated Hippocampal volumes}
  \label{SZ_volumes}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Screening Dataset Demographics Table                                  %
%                                                                             %
<<ADNI1-scr-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID < 2000,   
               method = "reverse", test=FALSE, overall=TRUE)
latex(tab, file="", size="scriptsize", 
    caption="ADNI1 1.5T Screening dataset demographics",
    title="tab:ADN1_1.5T_Screening_Dataset_Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c("295","413","619","685","729","782","938","954",
                 "1018","1155","907","981","222","324","448","546",
                 "553","572","602","610","814","1341","675","681",
                 "731","1130","41","68","70","101","249","293","316",
                 "414","698","1206","1222","1339","751","1030","3",
                 "5","10","16","22","53","168","183","241","1205",
                 "328","1095","991","213","159","337","343","642",
                 "753","1109","1217","29","56","1188","1293","149",
                 "382","431","1202")
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               adni_69_RIDs, method = "reverse", test=FALSE, overall=TRUE)

@@

<<ADN1-Xval_prep, include=FALSE>>=

# all data from the validation trials
t_data       = read.csv(gzfile('data/a2a_ants/results_2013_01_03.csv.gz'))
t_data       = subset(t_data, atlases %in% c(1,3,5,7,9))
a_data       = t_data[,] #read.csv(gzfile('data/a2a_tracc/results_2013_01_03.csv.gz'))
diagnoses    = read.csv('data/a2a_diagnoses.csv')
t_data$se    = NULL
t_data$sn    = NULL
t_data$j     = NULL
a_data$se    = NULL
a_data$sn    = NULL
a_data$j     = NULL       
# add registration method to data frames (before joining)
t_data       = cbind(t_data, reg_method = "ANIMAL")
a_data       = cbind(a_data, reg_method = "ANTS")

@@

all_data = merge(ma, mb, by=c("timestamp", "atlases", "batch", "label", "reg_method", "subject"))
all_data$k_diff = all_data$k.y - all_data$k.x 
method_labels <- list("Majority Vote" = "majvote", 
                      "NMI Vote" = "nmi", 
                      "Cross-correlation Vote" = "xcorr")
levels(all_data$method.y) <- method_labels
all_data       = merge(all_data, diagnoses)
remove(t_ma, a_ma, t_mb, a_mb, ma, mb)
@

\begin{figure}
<<ADN1-Xval-k_mean, cache=TRUE>>=
ggplot(all_data, aes(x=templates.y,y=k.y, colour=as.factor(atlases))) + 
  facet_grid(reg_method~method.y) + 
  #stat_summary(fun.y=mean,geom='line',
  #             aes(y=k.y)) + 
  stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "mean Kappa" )
@
  \caption{MAGeT brain mean kappa by atlas and template library size}
  \label{ADNI1_xval}
\end{figure}

\begin{figure}
<<ADN1-Xval-k_diff, cache=TRUE>>=

ggplot(subset(all_data, reg_method=="ANTS" & method.y=="Majority Vote"), 
       aes(x=templates.y, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  #stat_summary(fun.y=mean,geom='line',position="dodge",
  #             aes(y=k_diff,colour=as.factor(atlases))) + 
  stat_smooth(method=lm, formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Increase in mean Kappa between MAGeT brain and multiatlas" ) 
  #coord_cartesian(ylim=c(0,0.07), xlim=c(0.5,10.5))
@
  \caption{Increase in mean Kappa between MAGeT brain and multi-atlas}
  \label{ADNI1_cross_validation_MA_difference}
\end{figure}

\begin{figure}
<<ADN1-Xval-k_diff-ANTS-20tmpls>>=
a_20_tmpls = subset(all_data, 
                      reg_method  == "ANTS" &
                      method.y    == "Majority Vote" &
                      templates.y == 20)
ggplot(a_20_tmpls, aes(x=factor(atlases))) +
  stat_summary(fun.y=mean,geom='point',size=3,colour="red",aes(y=k.x)) +
  stat_summary(fun.y=mean,geom='line',colour="red",aes(y=k.x,group=1)) +
  stat_summary(fun.y=mean,geom='point',size=3,colour="blue",aes(y=k.y)) +
  stat_summary(fun.y=mean,geom='line',colour="blue",aes(y=k.y,group=1)) +
  labs(x="Number of Atlases", y = "Kappa", 
       title="mean Kappa of MA and MAGeT brain (ANTS/20 templates) \nvs\natlas library size") +
  scale_colour_hue(name="Method") + 
  coord_cartesian(ylim=c(0.7,0.9))
  #theme(legend.justification=c(1,0), legend.position=c(1,0))
@
  \caption{Change in mean Kappa between multi-atlas and MAGeT brain using ANTS, 20 templates}
  \label{}
\end{figure}

\begin{figure}
<<ADN1-Xval-k-by-disease, cache=TRUE>>=
a_20_tmpls = subset(all_data, 
                      reg_method  == "ANTS" &
                      method.y    == "Majority Vote" &
                      templates.y == 20 &
                      atlases %in% c(1,9))
template_library = read.csv('data/a2a_ants/template_library_composition.csv')
template_library$subject = NULL
a_20_tmpls = merge(a_20_tmpls, template_library,
                   by.x=c('timestamp', 'templates.y'), 
                   by.y=c('timestamp', 'templates'))
ggplot(a_20_tmpls, aes(y=k.y, x=templates_CN,colour=DX)) +  # why does k.x look the same?
  facet_grid( . ~ atlases) + 
  geom_smooth(method='lm',fullrange=F)

@@

<<ADNI-src-volumes-prep,echo=F,cache=T>>=
yr1   = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
mb    = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_MAGeT_volumes.csv")
maper = read.csv("data/ADNI_baseline_volumes/MAPER_volumes.csv")
snt   = read.csv("data/ADNI_baseline_volumes/UCSFSNTVOL.csv")
ucsd  = read.csv("data/ADNI_baseline_volumes/UCSDVOL.csv")
fs    = read.csv("data/ADNI_baseline_volumes/UCSFFSX_02_15_12.csv")
fsl   = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_FSLFIRST_volumes.csv", sep=';')

@@

\begin{figure}[h]
<<ADNI-src-volumes-boxplot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes-prep',cache=T>>=
totals = vols
qplot(DX,value,data=melt(totals,measure.vars=c("FS", "FSL","UCSD", "MAPER","MAGeT","SNT"),na.rm=T),
      colour=variable,geom="boxplot") + 
        xlab("Diagnosis") + ylab("Total Hippocampal Volume (mm3)")

@@

\begin{figure}[h]
<<ADNI-src-volumes-plot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes-boxplot',cache=T>>=
by_snt_volumes = melt(totals, measure.vars=c("FSL","MAPER","MAGeT", "FS", "UCSD"))
fsl=subset(by_snt_volumes, variable=="FSL")
maper=subset(by_snt_volumes, variable=="MAPER")
maget=subset(by_snt_volumes, variable=="MAGeT")
with(by_snt_volumes,
qplot(SNT,value,data=by_snt_volumes,
      colour=variable, geom="smooth", method="lm") + 
    geom_point() +
     geom_text(aes(
       label=paste("FSL       R=", round(cor(fsl$SNT,fsl$value, use="complete.obs"), 2), 
                   ", p<", round(as.double(t.test(fsl$SNT,fsl$value, paired=T)["p.value"]), 2)),
       x=1800,y=10000, hjust=0,size=2), data=data.frame()) + 
#     geom_text(aes(
#       label=paste("MAPER R=", round(cor(maper$SNT,maper$value, use="complete.obs"), 2), 
#                   ", p<", t.test(maper$SNT,maper$value, paired=T)["p.value"]), 
#       x=1800,y= 9500, hjust=0, size=2)) +
#     geom_text(aes(
#       label=paste("MAGeT  R=", round(cor(maget$SNT,maget$value, use="complete.obs"),2), 
#                   ", p<", t.test(maget$SNT,maget$value, paired=T)["p.value"]), 
#       x=1800,y= 9000, hjust=0, size=2)) +
    xlab("Total Hippocampal Volume (Manual Segmentation)") + 
    ylab("Total Hippocampal Volume (Automatic Segmentation)"))
@
  \caption{{\bf ADNI Baseline cohort.} Comparison of HC volumes by FreeSurfer
  (FSF), MAGeT brain (MAGeT), MAPER, and manual (SNT).}
  \label{ADNI-baseline-volumes-plot}
\end{figure}

<<ADNI-scr-summary-table, echo=F, results="asis",cache=FALSE>>=
melted = melt(totals, measure.vars=c("FSL","MAPER","SNT", "FS", "UCSD"))
cor_by_dx    = by(melted[c("MAGeT", "value")], c(melted['variable'], melted['DX']), 
                  function(x) cor(x$MAGeT,x$value, use="pairwise.complete.obs"))
cor_overall  = by(melted[c("MAGeT", "value")], melted['variable'], 
                  function(x) cor(x$MAGeT,x$value, use="pairwise.complete.obs"))
cor_by_dx = t(rbind(cor_by_dx))
cor_overall = rbind(cor_overall)

vol_by_dx = t(rbind(by(totals$MAGeT, totals$DX, mean)))
colnames(vol_by_dx) <- "Mean Volume"
vol_overall = matrix(mean(totals$MAGeT),dimnames=list("Overall"))

dx      = cbind(vol_by_dx, cor_by_dx)
overall <- cbind(vol_overall, cor_overall)

latex(format.df(rbind(dx, overall), dec=2),
      n.cgroup=c(1,5),cgroup=c("","Volume Correlation"),
      n.rgroup=c(4,1), file="", size="scriptsize", )

@

@@

\begin{figure}
<<FEP_volumes,echo=F,fig=T, fig.align='center',cache=T>>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv")
with(SZ_volumes,
qplot(Manual,MAGeT,data=SZ_volumes) + geom_smooth(method="lm") + 
   geom_text(aes(
       label=paste("r = ", round(cor(Manual,MAGeT, use="complete.obs"), 2), 
                   ", p <", round(as.numeric(t.test(Manual, MAGeT, paired=T)["p.value"]), 6))),
       x=7000,y=10000, hjust=0,vjust=-0.5,size=5,data=data.frame()) +
    xlab("Total Hippocampal Volume (Manual Segmentation)") + 
    ylab("Total Hippocampal Volume (MAGeT Segmentation)")
)
@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total
  HC volumes for MAGeT against manually rated Hippocampal volumes}
  \label{SZ_volumes}
\end{figure}
